### 2023-12-11 main.tex
''
l

{\color{magenta} Cristian I think it is good to keep some of the following citations, especailly the one to Schutz...
 \cite{schutz2017kardar, crampe2016integrable, finn2018matrix, floreani2022switching, casini2023density, redig2022ergodic, redig2023equilibrium, chatterjee2023multi,  franceschini2022orthogonal,floreani2023non,carinci2023open}
See also \cite{crampe2016integrable, finn2018matrix} for the asymmetric setup.}

@@

\paragraph{Proof of Proposition~\ref{propositin-duality-RD}:}
we only prove duality for the bulk generator, since the boundary generators are the same of the multispecies stirring process. In the spirit of Section~\ref{sectionDuality}, it is convenient to write the Hamiltonian of the process, that is defined through the basis elements \eqref{actionE} of the ${gl}(N)$ Lie algebra as 
\begin{equation}
	H^{rd}=H_{\text{left}}+\sum_{x=1}^{L-1}\left(\sigma_{11}\mathcal{H}_{x,x+1}+\sigma_{12}\sum_{c=1}^{N-2}\mathcal{H}_{x,x+1}^{c}+(\Upsilon-2\sigma_{12})\mathcal{H}_{x,x+1}^{m}\right)+H_{\text{right}}\dt
\end{equation}
Here, $\mathcal{H}_{x,x+1}$ is  \eqref{edgeHamiltonian}  and $H_{\text{left}},H_{\text{right}}$ is \eqref{siteHamiltonian} (with the choice of the boundary parameters done in \eqref{boundaryParamRD}). We, then have that 
\begin{equation}
	\mathcal{H}_{x,x+1}^{c}=\sum_{A,B=1}^{N}\left(E_{h_{c}(A)B}\otimes E_{h_{c}(B)A}-E_{AA}\otimes E_{BB}\right)
\end{equation}
and 
\begin{equation}
	\mathcal{H}_{x,x+1}^{m}=\sum_{A,B=1}^{N}\left(E_{BA}\otimes \mathbbm{1}-E_{AA}\otimes \mathbbm{1}\right)\dt
\end{equation}
We prove that, for arbitrary $x\in\{1,\ldots,L\}$,
\begin{align}
	&(\mathcal{H}_{x,x+1}^{c})^{T}D=D\mathcal{H}_{x,x+1}^{c}\qquad \forall c\in \{1,\ldots,N-2\}\label{cDualityRelation}\\
	&(\mathcal{H}_{x,x+1}^{m})^{T}D=D\mathcal{H}_{x,x+1}^{m}\label{mDualityRelation}\dt
\end{align}
Since the matrix $R_{x}$ is the same as in Section~\ref{sectionDuality}, we shall show that $\mathcal{H}_{x,x+1}^{c}$ and $\mathcal{H}_{x,x+1}^{m}$ commute with $\exp{(E^{x})}\exp{(E^{x+1})}$, where $E$ is defined in \eqref{EquationEx}. Using \eqref{eq:comgl}, the bilinearity and associativity of the Kronecker product and the bilinearity of the brackets we obtain 
\begin{align*}
	&\left[\left(\sum_{a=1}^{N-1}E_{aN}^{x}\otimes\sum_{a=1}^{N-1}E_{aN}^{x+1}\right), \sum_{D,K=1}^{N}\left(E_{h_{c}(K)D}^{x}\otimes E_{h_{c}(D)K}^{x+1} -E_{DD}^{x}\otimes E_{KK}^{x+1}\right)\right]
	\\=&
%	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left((E_{da}^{x}\delta_{h_{c}(b)N}-E_{h_{c}(b)N}^{x}\delta_{ad})-(E_{da}^{x}\delta_{aN}-E_{aN}^{x}\delta_{da})\right)\right. \\ \otimes&\left. \left((E_{kb}^{x+1}\delta_{h_{c}(a)N}-E_{h_{c}N}^{x+1}\delta_{kb})-(E_{kb}^{x+1}\delta_{bN}-E_{bN}^{x+1}\delta_{kb})\right)\right\}
%	\\=&
%	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left(E_{da}^{x}-E_{h_{c}(b)N}^{x}-E_{da}^{x}+E_{aN}^{x}\right)\otimes\left(E_{kb}^{x+1}-E_{h_{c}(a)N}^{x+1}-E_{kb}^{x+1}+E_{bN}^{x+1}\right)\right\}
%	\\=&
	\sum_{a,b=1}^{N-1}\sum_{D,k=1}^{N}\left\{\left(E_{DN}^{x}-E_{h_{c}(k)N}^{x}\right)\otimes \left(E_{DN}^{x+1}-E_{h_{c}(K)N}^{x+1}\right)\right\}=0\dt
\end{align*}
In the last equality we used the fact that map $h_{c}(\cdot)$ is surjective. 
%\paragraph{Duality for $H_{x,x+1}^{m}$} 
Concerning the commutator between $\mathcal{H}_{x,x+1}^{m}$ the proof is similar and gives
\begin{equation}
	\left[\sum_{a=1}^{N-1}E_{aN}^{x}\otimes \sum_{b=1}^{N-1}E_{bN}^{x},\;\sum_{D,K=1}^{N}\left(E_{K a}^{x}\otimes \mathbbm{1}^{x+1}-E_{DD}^{x}\otimes \mathbbm{1}^{x+1}\right)\right]=0\dt
\end{equation}
\begin{flushright}
	$\square$
\end{flushright}

@@

\begin{equation}
	\mathcal{L}_{x,x+1}n_{d}^{x}=\sum_{A,B=1}^{N}n_{A}^{x}n_{B}^{x+1}\left((n_{d}^{x}-\bm{\delta}_{A}^{x}+\bm{\delta}_{B}^{x}+\bm{\delta}_{A}^{x+1}-\bm{\delta}_{B}^{x+1})-n_{d}^{x}\right)
\end{equation}
In the brackets of the right-hand-side of the above equation we have: a contribution $-n_{d}^{x}n_{B}^{x+1}$, when $A=d$; a contribution $n_{A}^{x}n_{d}^{x+1}$, when $B=d$. Thus we obtain 
\begin{align*}
	\mathcal{L}_{x,x+1}n_{d}^{x}=&\sum_{a=1}^{N}n_{A}^{x}n_{d}^{x+1}-\sum_{bB=1}^{N}n_{d}^{x}n_{B}^{x+1}\\
	=&\sum_{a=1}^{N-1}n_{a}^{x}n_{d}^{x+1}+(\nu-\sum_{i=1}^{N-1}n_{i}^{x})n_{d}^{x+1}-\sum_{b=1}^{N-1}n_{d}^{x}n_{b}^{x+1}-n_{d}^{x}(\nu-\sum_{i=1}^{N-1}n_{i}^{x+1})\\
	=&\nu\left(n_{d}^{x+1}-n_{d}^{x}\right)
\end{align*} 
where we have used the fact that $n_{N}^{x}=\nu-\sum_{i=1}^{N-1}n_{i}^{x}$.
\item \textit{Stirring-mutation generator}: for every $c\in \{1,\ldots,N-2\}$ we have 
\begin{equation}
	\begin{split}
		\mathcal{L}_{x,x+1}^{c}n_{d}^{x}&=\sum_{A,B=1}^{N}n_{A}^{x}n_{B}^{x+1}\left((n_{d}^{x}-\bm{\delta}_{A}^{x}+\bm{\delta}_{h_{c}(B)}^{x}+\bm{\delta}_{h_{c}(A)}^{x+1}-\bm{\delta}_{B}^{x+1})-n_{d}^{x}\right)
	\end{split}
\end{equation}
In the brackets of the right-hand-side of the above equation we have: a contribution $-n_{d}^{x}n_{K}^{x+1}$, when $A=d$ (here we call $K=h_{c}(B)$); a contribution $n_{A}^{x}n_{q}^{x+1}$, when $h_{c}(B)=d$. Thus we obtain 
\begin{align}
		\mathcal{L}_{x,x+1}^{c}n_{d}^{x}&=\sum_{A=1}^{N}n_{A}^{x}n_{b}^{x+1}-\sum_{K=1}^{N}n_{d}^{x}n_{K}^{x+1}
		\\&=
		\sum_{k=1}^{N-1}n_{k}^{x}n_{b}^{x+1}+\left(\nu-\sum_{i=1}^{N-1}n_{i}^{x}\right)n_{b}^{x+1}-\sum_{K=1}^{N-1}n_{d}^{x}n_{K}^{x+1}-n_{d}^{x}\left(\nu-\sum_{i=1}^{N-1}n_{i}^{x+1}\right)
		\\&=
		\nu(n_{b}^{x+1}-n_{d}^{x})
\end{align}
where  we used the fact that $n_{N}^{x}=\nu-\sum_{i=1}^{N-1}n_{i}^{x}$. Let us observe that, by the definition of the map $h_{c}(\cdot)$ and for the fact that $1\leq c\leq N-2$, we have that in the above equation $B\neq d$ and $B\neq N$.
\item \textit{Pure-mutation generator}: here we have that 
\begin{equation}
	\mathcal{L}_{x,y}^{m}n_{d}^{x}=\sum_{A,B=1}^{N-1}n_{A}^{x}\left((n_{d}^{x}-\bm{\delta}_{A}^{x}+\bm{\delta}_{B}^{x})-n_{d}^{x}\right)=\sum_{b=1}^{N-1}(n_{b}^{x}-n_{d}^{x})
\end{equation}
\end{enumerate}
Then, summing over all indices $c\in\{1,\ldots,N-2\}$, we obtain \eqref{actionGraphRD} and \eqref{DifferenceEquation} follows.
\paragraph{Action of $\mathcal{L}_{\text{left}}^{rd}$ on the occupation variable}
Acting on $n_{d}^{1}$  with $\mathcal{L}_{\text{left}}^{rd}$ we obtain
\begin{equation}\label{actionBoudaryRD}
	\begin{split}
		\mathcal{L}_{\text{left}}^{rd}n_{d}^{1}&=\sum_{A,B=1}^{N}\alpha_{A}n_{B}^{1}\left((n_{d}^{1}-\bm{\delta}_{B}^{1}+\bm{\delta}_{B}^{1})-n_{d}^{1}\right)=\alpha_{d}\sum_{B=1\,:\,b\neq d}^{N}n_{B}^{1}-n_{d}^{1}\sum_{B=1\,:B\neq d}^{N}\alpha_{B}
	\end{split}
\end{equation}
Using equation \eqref{boundaryParamRD}, we obtain 
\begin{align}
			\mathcal{L}_{\text{left}}^{rd}n_{d}^{1}&=
		\alpha_{d}\left(\nu-n_{d}^{1}\right)-n_{d}^{1}\left(\nu(\sigma_{11}+(N-2)\sigma_{12})-\alpha_{d}\right)
		\\&=\alpha_{d}\nu-n_{d}^{1}\nu(\sigma_{11}+(N-2)\sigma_{12})
		\\&=
		\nu\sigma_{11}\rho_{d}^{\text{left}}+\nu\sigma_{12}\left(\sum_{b=1\,:\,b\neq d}^{N-1}\rho_{b}^{\text{left}}\right)-n_{d}^{1}\nu(\sigma_{11}+(N-2)\sigma_{12})
		\\&=
		\nu \sigma_{11} (\rho_{d}^{\text{left}}-n_{d}^{1})+\nu\sigma_{12}\sum_{b=1\,:\,b\neq d}^{N-1}(\rho_{b}^{\text{left}}-n_{d}^{1})
	\end{align}
Therefore, \eqref{actionBDLine} follows.
\bibliography{ref}
\bibliographystyle{unsrt}

>>>>

{\color{magenta} Cristian I think it is good to keep some of the following citations, especailly the one to Schutz...
 \cite{schutz2017kardar, crampe2016integrable, finn2018matrix, chatterjee2023multi,redig2022ergodic, floreani2022switching} and \cite{casini2023density,  franceschini2022orthogonal,redig2023equilibrium,floreani2023non,carinci2023open}
See also \cite{crampe2016integrable, finn2018matrix} for the asymmetric setup.}

@@

\paragraph{Proof of Proposition~\ref{propositin-duality-RD}:}
we only prove duality for the bulk generator, since the boundary generators are the same of the multispecies stirring process. In the spirit of Section~\ref{sectionDuality}, it is convenient to write the Hamiltonian of the process, that is defined through the basis elements \eqref{actionE} of the ${gl}(N)$ Lie algebra as 
\begin{equation}
	H^{rd}=H_{\text{left}}+\sum_{x=1}^{L-1}\left(\sigma_{11}\mathcal{H}_{x,x+1}+\sigma_{12}\sum_{c=1}^{N-2}\mathcal{H}_{x,x+1}^{c}+(\Upsilon-2\sigma_{12})\mathcal{H}_{x,x+1}^{m}\right)+H_{\text{right}}\dt
\end{equation}
Here, $\mathcal{H}_{x,x+1}$ is  \eqref{edgeHamiltonian}  and $H_{\text{left}},H_{\text{right}}$ is \eqref{siteHamiltonian} (with the choice of the boundary parameters done in \eqref{boundaryParamRD}). We, then have that 
\begin{equation}
	\mathcal{H}_{x,x+1}^{c}=\sum_{A,B=1}^{N}\left(E_{h_{c}(A)B}\otimes E_{h_{c}(B)A}-E_{AA}\otimes E_{BB}\right)
\end{equation}
and 
\begin{equation}
	\mathcal{H}_{x,x+1}^{m}=\sum_{A,B=1}^{N}\left(E_{BA}\otimes \mathbbm{1}-E_{AA}\otimes \mathbbm{1}\right)\dt
\end{equation}
We prove that, for arbitrary $x\in\{1,\ldots,L\}$,
\begin{align}
	&(\mathcal{H}_{x,x+1}^{c})^{T}D=D\mathcal{H}_{x,x+1}^{c}\qquad \forall c\in \{1,\ldots,N-2\}\label{cDualityRelation}\\
	&(\mathcal{H}_{x,x+1}^{m})^{T}D=D\mathcal{H}_{x,x+1}^{m}\label{mDualityRelation}\dt
\end{align}
Since the matrix $R_{x}$ is the same as in Section~\ref{sectionDuality}, we shall show that $\mathcal{H}_{x,x+1}^{c}$ and $\mathcal{H}_{x,x+1}^{m}$ commute with $\exp{(E^{x})}\exp{(E^{x+1})}$, where $E$ is defined in \eqref{EquationEx}. Using \eqref{eq:comgl}, the bilinearity and associativity of the Kronecker product and the bilinearity of the brackets we obtain 
\begin{align*}
	&\left[\left(\sum_{a=1}^{N-1}E_{aN}^{x}\otimes\sum_{a=1}^{N-1}E_{aN}^{x+1}\right), \sum_{D,K=1}^{N}\left(E_{h_{c}(K)D}^{x}\otimes E_{h_{c}(D)K}^{x+1} -E_{DD}^{x}\otimes E_{KK}^{x+1}\right)\right]
	\\=&
%	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left((E_{da}^{x}\delta_{h_{c}(b)N}-E_{h_{c}(b)N}^{x}\delta_{ad})-(E_{da}^{x}\delta_{aN}-E_{aN}^{x}\delta_{da})\right)\right. \\ \otimes&\left. \left((E_{kb}^{x+1}\delta_{h_{c}(a)N}-E_{h_{c}N}^{x+1}\delta_{kb})-(E_{kb}^{x+1}\delta_{bN}-E_{bN}^{x+1}\delta_{kb})\right)\right\}
%	\\=&
%	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left(E_{da}^{x}-E_{h_{c}(b)N}^{x}-E_{da}^{x}+E_{aN}^{x}\right)\otimes\left(E_{kb}^{x+1}-E_{h_{c}(a)N}^{x+1}-E_{kb}^{x+1}+E_{bN}^{x+1}\right)\right\}
%	\\=&
	\sum_{a,b=1}^{N-1}\sum_{D,K=1}^{N}\left\{\left(E_{DN}^{x}-E_{h_{c}(K)N}^{x}\right)\otimes \left(E_{DN}^{x+1}-E_{h_{c}(K)N}^{x+1}\right)\right\}=0\dt
\end{align*}
In the last equality we used the fact that map $h_{c}(\cdot)$ is surjective. 
%\paragraph{Duality for $H_{x,x+1}^{m}$} 
Concerning the commutator between $\mathcal{H}_{x,x+1}^{m}$ the proof is similar and gives
\begin{equation}
	\left[\sum_{a=1}^{N-1}E_{aN}^{x}\otimes \sum_{b=1}^{N-1}E_{bN}^{x},\;\sum_{D,K=1}^{N}\left(E_{K a}^{x}\otimes \mathbbm{1}^{x+1}-E_{DD}^{x}\otimes \mathbbm{1}^{x+1}\right)\right]=0\dt
\end{equation}
\begin{flushright}
	$\square$
\end{flushright}

@@

\begin{equation}
	\mathcal{L}_{x,x+1}n_{d}^{x}=\sum_{A,B=1}^{N}n_{A}^{x}n_{B}^{x+1}\left((n_{d}^{x}-\bm{\delta}_{A}^{x}+\bm{\delta}_{B}^{x}+\bm{\delta}_{A}^{x+1}-\bm{\delta}_{B}^{x+1})-n_{d}^{x}\right)
\end{equation}
In the brackets of the right-hand-side of the above equation we have: a contribution $-n_{d}^{x}n_{B}^{x+1}$, when $A=d$; a contribution $n_{A}^{x}n_{d}^{x+1}$, when $B=d$. Thus we obtain 
\begin{align*}
	\mathcal{L}_{x,x+1}n_{d}^{x}=&\sum_{a=1}^{N}n_{A}^{x}n_{d}^{x+1}-\sum_{B=1}^{N}n_{d}^{x}n_{B}^{x+1}\\
	=&\sum_{a=1}^{N-1}n_{a}^{x}n_{d}^{x+1}+(\nu-\sum_{i=1}^{N-1}n_{i}^{x})n_{d}^{x+1}-\sum_{b=1}^{N-1}n_{d}^{x}n_{b}^{x+1}-n_{d}^{x}(\nu-\sum_{i=1}^{N-1}n_{i}^{x+1})\\
	=&\nu\left(n_{d}^{x+1}-n_{d}^{x}\right)
\end{align*} 
where we have used the fact that $n_{N}^{x}=\nu-\sum_{i=1}^{N-1}n_{i}^{x}$.
\item \textit{Stirring-mutation generator}: for every $c\in \{1,\ldots,N-2\}$ we have 
\begin{equation}
	\begin{split}
		\mathcal{L}_{x,x+1}^{c}n_{d}^{x}&=\sum_{A,B=1}^{N}n_{A}^{x}n_{B}^{x+1}\left((n_{d}^{x}-\bm{\delta}_{A}^{x}+\bm{\delta}_{h_{c}(B)}^{x}+\bm{\delta}_{h_{c}(A)}^{x+1}-\bm{\delta}_{B}^{x+1})-n_{d}^{x}\right)
	\end{split}
\end{equation}
In the brackets of the right-hand-side of the above equation we have: a contribution $-n_{d}^{x}n_{K}^{x+1}$, when $A=d$ (here we call $K=h_{c}(B)$); a contribution $n_{A}^{x}n_{q}^{x+1}$, when $h_{c}(B)=d$. Thus we obtain 
\begin{align}
		\mathcal{L}_{x,x+1}^{c}n_{d}^{x}&=\sum_{A=1}^{N}n_{A}^{x}n_{b}^{x+1}-\sum_{K=1}^{N}n_{d}^{x}n_{K}^{x+1}
		\\&=
		\sum_{k=1}^{N-1}n_{k}^{x}n_{b}^{x+1}+\left(\nu-\sum_{i=1}^{N-1}n_{i}^{x}\right)n_{b}^{x+1}-\sum_{K=1}^{N-1}n_{d}^{x}n_{K}^{x+1}-n_{d}^{x}\left(\nu-\sum_{i=1}^{N-1}n_{i}^{x+1}\right)
		\\&=
		\nu(n_{b}^{x+1}-n_{d}^{x})
\end{align}
where  we used the fact that $n_{N}^{x}=\nu-\sum_{i=1}^{N-1}n_{i}^{x}$. Let us observe that, by the definition of the map $h_{c}(\cdot)$ and for the fact that $1\leq c\leq N-2$, we have that in the above equation $B\neq d$ and $B\neq N$.
\item \textit{Pure-mutation generator}: here we have that 
\begin{equation}
	\mathcal{L}_{x,y}^{m}n_{d}^{x}=\sum_{A,B=1}^{N-1}n_{A}^{x}\left((n_{d}^{x}-\bm{\delta}_{A}^{x}+\bm{\delta}_{B}^{x})-n_{d}^{x}\right)=\sum_{b=1}^{N-1}(n_{b}^{x}-n_{d}^{x})
\end{equation}
\end{enumerate}
Then, summing over all indices $c\in\{1,\ldots,N-2\}$, we obtain \eqref{actionGraphRD} and \eqref{DifferenceEquation} follows.
\paragraph{Action of $\mathcal{L}_{\text{left}}^{rd}$ on the occupation variable}
Acting on $n_{d}^{1}$  with $\mathcal{L}_{\text{left}}^{rd}$ we obtain
\begin{equation}\label{actionBoudaryRD}
	\begin{split}
		\mathcal{L}_{\text{left}}^{rd}n_{d}^{1}&=\sum_{A,B=1}^{N}\alpha_{A}n_{B}^{1}\left((n_{d}^{1}-\bm{\delta}_{B}^{1}+\bm{\delta}_{B}^{1})-n_{d}^{1}\right)=\alpha_{d}\sum_{B=1\,:\,B\neq d}^{N}n_{B}^{1}-n_{d}^{1}\sum_{B=1\,:B\neq d}^{N}\alpha_{B}
	\end{split}
\end{equation}
Using equation \eqref{boundaryParamRD}, we obtain 
\begin{align}
			\mathcal{L}_{\text{left}}^{rd}n_{d}^{1}&=
		\alpha_{d}\left(\nu-n_{d}^{1}\right)-n_{d}^{1}\left(\nu(\sigma_{11}+(N-2)\sigma_{12})-\alpha_{d}\right)
		\\&=\alpha_{d}\nu-n_{d}^{1}\nu(\sigma_{11}+(N-2)\sigma_{12})
		\\&=
		\nu\sigma_{11}\rho_{d}^{\text{left}}+\nu\sigma_{12}\left(\sum_{b=1\,:\,b\neq d}^{N-1}\rho_{b}^{\text{left}}\right)-n_{d}^{1}\nu(\sigma_{11}+(N-2)\sigma_{12})
		\\&=
		\nu \sigma_{11} (\rho_{d}^{\text{left}}-n_{d}^{1})+\nu\sigma_{12}\sum_{b=1\,:\,b\neq d}^{N-1}(\rho_{b}^{\text{left}}-n_{d}^{1})
	\end{align}
Therefore, \eqref{actionBDLine} follows.
\bibliography{ref}
\bibliographystyle{unsrt}
